---
title: "Segmenting histone-mark signal"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Segmenting histone-mark signal}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

```{r setup}
library(chiptools)
```

The central chromatin problem in this analysis: given normalised ChIP-seq
signal at a few thousand promoters, decide for each gene and each mark
whether the mark is **absent**, **present**, or **present at an extreme level
that is probably not the same biological thing**.

A single threshold does not work, for two reasons visible in the data. The
distribution is bimodal — background and real signal — so the sensible cut
between them is where the two components cross, not at a round number. And
there is a long right tail of a few hundred very highly bound promoters that,
left in, drags the signal component upward and moves that crossing point.

So the segmentation is three thresholds, found in two steps.

```{r, eval = FALSE}
# 1. where does the upper tail start?
cutoffs <- trim_upper_tail(histone_rlog_prune, derivative_threshold = 4e-4)

# 2. mixture threshold per sample, fitted with the tail excluded
breaks <- chip_binarize_tail_removed(histone_rlog_prune, cutoffs)

# 3. call every gene, then collapse replicates within condition
calls <- chip_categorize_and_collapse(signal_with_gene_id, breaks)
```

## Step 1: find the tail

`trim_upper_tail()` ranks each sample's values, LOESS-smooths the rank/value
curve, differentiates it, and takes the knee — the rank at which the smoothed
first derivative falls below `derivative_threshold`. Below that rank, signal
climbs steeply with rank; above it, you are in the tail.

```{r}
set.seed(1)
# background, real signal, and a small extreme tail
signal <- c(stats::rnorm(1500, 0.25, 0.06),
            stats::rnorm(900,  0.62, 0.09),
            stats::rnorm(120,  0.95, 0.03))
dat <- data.frame(sample_a = signal)

hist(dat$sample_a, breaks = 80, col = "grey", border = "white",
     main = "Two components plus a tail", xlab = "normalised signal")
```

The threshold is **on the scale of the smoothed derivative**, which depends on
how many rows you have. It does not transfer between data sets. The notebooks
used `4e-4` for the GNP promoter set and `5.5e-4` for the MB set, and both
were chosen by looking at the lower panel of the diagnostic plot
`trim_upper_tail()` prints. Treat any default as a starting point:

```{r, eval = FALSE}
# print the diagnostics, decide, then re-run
cutoffs <- trim_upper_tail(histone_rlog_prune, derivative_threshold = 4e-4,
                           generate_plots = TRUE)
cutoffs
#>          Column CrossingRank CrossingSmoothedValue
#> 1      sample_a         2402             0.8161...
```

Samples where no crossing is found are dropped from the table, so check
`nrow(cutoffs)` against `ncol(histone_rlog_prune)`.

## Step 2: the bound/unbound threshold

With the tail known, `chip_binarize_tail_removed()` runs
`binarize_counts()` per sample on the values *below* that sample's tail
cutoff. `binarize_counts()` fits a two-component normal mixture and solves
for the point where the posterior probability of belonging to the background
component equals 0.5:

```{r}
if (requireNamespace("mixtools", quietly = TRUE)) {
  below_tail <- signal[signal < 0.88]
  fit <- binarize_counts(below_tail, plot = FALSE, return_what = "summary")
  round(fit, 4)
}
```

`cutoff_50` is the posterior-0.5 crossing and is the bound/unbound call.
`cutoff_75` is the stricter 0.75 crossing, and is what the returned `bin`
column actually uses — worth knowing, since it means the default 0/1 call is
conservative.

Fitting below the tail matters. Here is the same fit with the tail left in:

```{r}
if (requireNamespace("mixtools", quietly = TRUE)) {
  with_tail <- binarize_counts(signal, plot = FALSE, return_what = "summary")
  without   <- binarize_counts(signal[signal < 0.88], plot = FALSE,
                               return_what = "summary")
  data.frame(
    row.names  = c("tail included", "tail excluded"),
    cutoff_50  = c(with_tail[["cutoff_50"]],  without[["cutoff_50"]]),
    mean_upper = c(with_tail[["mean_upper"]], without[["mean_upper"]])
  )
}
```

The signal component's mean is pulled up by the tail, and the threshold moves
with it.

### If the mixture will not fit

- **A dominant zero bin.** Count data over promoters often has a zero spike
  tall enough to swamp every other bin and capture the whole background
  component. `soften_zero()` replaces a random subset of the zeros with `NA`
  so the spike is at most `scale_max` times the tallest other bin.
- **Right-skewed, un-logged signal.** A gamma mixture fits better than a
  normal one: `binarize_counts_p()`, which estimates its own starting values
  from the data quantiles and retries with jitter.
- **More than two states.** `trinarize_counts()` for three, or
  `split_counts()` to let BIC choose the component count.

## Step 3: call genes and collapse replicates

`chip_categorize_and_collapse()` applies both thresholds per replicate to get
`unbound` / `bound` / `tail`, then collapses the replicates of each condition
with a precedence rule — **tail beats bound, bound beats unbound**. A gene
called `tail` in any replicate is `tail` for that condition.

```{r, eval = FALSE}
calls <- chip_categorize_and_collapse(signal_with_gene_id, breaks)
table(calls$condition, calls$category)
```

The input needs a `gene_id` column and one column per replicate; `breaks`
needs `replicate_names`, `condition`, `cutoff_50` and
`CrossingSmoothedValue`. In practice `breaks` is the `chip_binarize_tail_removed()`
output left-joined to a replicate-to-condition table.

## The clustering alternative

Instead of thresholds, cluster the values. `chip_mclust_icl_all()` fits
Gaussian mixtures per sample over a range of component counts, picks one by
ICL, and gives the held-out tail its own top class:

```{r, eval = FALSE}
classes <- chip_mclust_icl_all(histone_rlog_prune, cutoffs, G_range = 2:8)
```

`chip_segment()` is `trim_upper_tail()` then `chip_mclust_icl_all()` in one
call:

```{r, eval = FALSE}
seg <- chip_segment(histone_rlog_prune, deriv_cutoff = 4e-4)
seg$cutoffs
head(seg$classes)
```

Use the threshold route when you want a call you can explain in a figure
legend, and the clustering route when you want to discover how many states
there are. The notebooks use thresholds for the main figures.

## Normalisation upstream

Both routes assume the samples are on a comparable scale.
`wins_norm_histones()` clips each column at a fixed *number of points* from
each end — not a quantile — then rescales to 0–1:

```{r, eval = FALSE}
histone_norm <- wins_norm_histones(histone_rlog_bm, low = 500, high = 20)
```

Clipping by point count keeps the rescaling stable when one sample has a
handful of extreme outliers, which quantile winsorising does not. Note that
`median_subtract` cannot change the default output: min-max rescaling is
shift-invariant, so a constant shift cancels. It is observable only with
`normalize = FALSE`. See `?wins_norm_histones`.

`plot_norm_heatmap()` is the other normalisation, for plotting rather than
calling: divide each gene by its own row mean and take `log2`, giving a
matrix symmetric about zero that a diverging palette reads correctly.

## Downstream

Once genes are called, the comparisons:

- `chip_scatter_cutoffs()` — two samples scattered with the bound-not-tail
  window drawn as a box, a gene set highlighted, marginal densities
- `chip_gene_group_comp()` — binding in early vs late genes, with a Wilcoxon
  p-value. Pass a `t50_cutoff`; with `NULL` it filters `t50_cat` to
  `"Late"`/`"Mid"` and then tests for `"early"`/`"late"`, so the p-value is
  `NA` by construction
- `plot_tsne_kmeans()` — t-SNE of the signal matrix beside a cluster-ordered
  heatmap, colours matched between panels by `opti_map()`
- `chip_de_rpm()` — DESeq2 differential binding between two conditions
