---
title: "Time courses, t50, and gene classification"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Time courses, t50, and gene classification}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

```{r setup}
library(chiptools)
```

The differentiation programme is a *sequence*: genes turn on in an order. To
compare that order against chromatin, each gene needs a single number saying
when it turns on. That number is the **t50** — the time at which expression
reaches half its range.

Getting there means reconciling two studies that sampled cerebellar
development at different ages.

```{r, eval = FALSE}
interp  <- time_expr_interp(expr_list, time_ref)
aligned <- time_expr_align(interp)
pooled  <- time_expr_avg(aligned)
t50     <- expr_half_max(pooled)
```

## The time reference

`time_ref` is the whole construction. One row per age on the full axis, one
logical column per study marking where that study has a sample:

```{r}
time_ref <- data.frame(time = c(paste0("E", 15:20), paste0("P", 0:56)))
time_ref$scott  <- time_ref$time %in% c("E15", "P1", "P7", "P14")
time_ref$hatten <- time_ref$time %in% c("P0", "P7", "P12", "P18", "P21", "P56")
rownames(time_ref) <- time_ref$time

head(time_ref, 10)
sum(time_ref$scott)
sum(time_ref$hatten)
```

Column order must match the order of `expr_list`, and each study's expression
matrix must have its columns in time order. Nothing checks this for you, and
getting it wrong produces a plausible-looking wrong answer — so it is worth
asserting:

```{r, eval = FALSE}
stopifnot(
  ncol(expr_list$scott)  == sum(time_ref$scott),
  ncol(expr_list$hatten) == sum(time_ref$hatten),
  identical(names(expr_list), c("scott", "hatten"))
)
```

## Interpolate, align, pool

`time_expr_interp()` linearly interpolates each study onto the full axis,
giving matrices of the same shape with `NA` outside each study's sampled
range:

```{r}
# a toy gene measured by both studies, on their own schedules
scott  <- matrix(c(1, 3, 7, 9), nrow = 1,
                 dimnames = list("Gene1", NULL))
hatten <- matrix(c(2.5, 5.5, 7.5, 8.5, 9, 9.5), nrow = 1,
                 dimnames = list("Gene1", NULL))

interp <- time_expr_interp(list(scott = scott, hatten = hatten),
                           time_ref[, c("scott", "hatten")])

length(interp)
dim(interp[[1]])
round(interp[[1]][1, 1:12], 2)
```

`time_expr_align()` then removes the per-gene offset between studies, using
the ages where they overlap. Without it, pooling averages a systematic
platform difference — TRAP microarray against RNA-seq — into the trajectory,
which shifts t50. `time_expr_avg()` pools the aligned matrices element-wise,
ignoring `NA`:

```{r}
pooled <- time_expr_avg(interp)
dim(pooled)
round(pooled[1, 1:12], 2)
```

`time_expr_align()` is written for exactly two studies; the source carries
`2DS FIX` markers where that assumption is baked in.

## t50

```{r}
t50 <- expr_half_max(pooled)
t50
```

`expr_half_max()` finds the midpoint between each gene's minimum and maximum
and interpolates the crossing, restricted to the monotonic stretch between
those two extremes — so a gene that rises, falls, and rises again gets the
timing of its dominant move rather than an average of all three.

Genes whose maximum is below zero return `NA`, and the result is capped at
the last genuinely sampled time point.

### Genes that go up and then down

For those, one t50 is misleading. `expr_half_max_min()` splits each
trajectory into its dominant move and the largest opposing move elsewhere,
reporting a timing, a log2 fold change and a direction for each:

```{r}
up_then_down <- matrix(c(1, 4, 8, 9, 7, 3, 1), nrow = 1,
                       dimnames = list("Gene2", NULL))
expr_half_max_min(up_then_down)
```

A zero means no move was found in that direction. **This works on a matrix,
not a list** — the list branch collapses the six-element result to a single
number. Use `lapply()` over your studies.

### Describing the shape

`time_of_change()` gives a fuller description: where each trajectory rises and
falls, for how long, and how much of the total change each run accounts for.

```{r}
tc <- time_of_change(rbind(Gene1 = pooled[1, 1:20],
                           Gene2 = rev(pooled[1, 1:20])), thres = 0.2)
tc[, c("max.val", "min.val", "change.total", "inc.span", "dec.span")]
```

`inc.start` and `inc.stop` use the sentinels 100 and -100 when no run exists,
not `NA`.

### Back onto real ages

t50 comes out on the dense interpolated index, which is awkward to plot
against six sampling ages. `t50_to_compressed()` maps it onto the compressed
axis of actually-sampled points:

```{r}
t50_to_compressed(t50, time_ref[, c("scott", "hatten")])
```

## Classifying genes

With a t50 and a time of maximum per gene, `classify_timecourse_genes()` cuts
genes into developmental windows on both axes, then overrides genes that do
not change or are not expressed:

```{r}
gene_table <- data.frame(
  row.names = c("early", "mid", "late", "flat", "silent"),
  avg_max = c(1, 5, 30, 10, 1),
  avg_t50 = c(2, 10, 25, 20, 2),
  avg_fc  = c(3, 4, 5, 0.2, 3),
  # "mid" is significant in scott only -- that is what pval_both decides on
  hatten_trap_min_pval = c(1e-4, 0.50, 1e-4, 0.5, 1e-4),
  scott_gnp_min_pval   = c(1e-4, 1e-4, 1e-4, 0.5, 1e-4),
  max_expr_Scott  = c(500, 500, 500, 500, 2),
  max_expr_Hatten = c(500, 500, 500, 500, 2)
)

classified <- classify_timecourse_genes(
  gene_table,
  fc_cutoff    = 2,
  p_val_cutoff = 0.001,
  expr_cutoffs = c(max_expr_Scott = 100, max_expr_Hatten = 100)
)

classified[, c("max_cat", "t50_cat", "max_t50_cat")]
```

Three things about the overrides are deliberate and worth knowing:

**Order.** `"unexpressed"` is applied *after* `"unchanged"`, so a gene that is
both comes back as unexpressed.

**`"unexpressed"` needs every floor.** A gene must be below the expression
cutoff in *all* data sets. One study detecting it is enough to keep it.

**`pval_both` is the consequential switch.** `FALSE` (the default) means a
gene need clear the p-value threshold in only one study; `TRUE` means both.
Here `"mid"` clears the threshold in scott but not hatten, so it survives as
a real gene under `FALSE` and is written off as `"unchanged"` under `TRUE`.
The notebooks ran their cutoff sweeps with it set both ways:

```{r}
lenient <- classify_timecourse_genes(gene_table, 2, 0.001,
                                     c(max_expr_Scott = 100,
                                       max_expr_Hatten = 100),
                                     pval_both = FALSE, table_only = TRUE)
strict  <- classify_timecourse_genes(gene_table, 2, 0.001,
                                     c(max_expr_Scott = 100,
                                       max_expr_Hatten = 100),
                                     pval_both = TRUE, table_only = TRUE)
rbind(lenient = lenient, strict = strict)[, c("Max_P0_7 t50_Mid",
                                              "unchanged", "unexpressed")]
```

`table_only = TRUE` returns counts rather than the data frame, and because
the override levels are declared up front, every call returns the same 11
named entries including the empty ones — so a sweep `rbind`s cleanly.

## Binning along t50

For enrichment *along* the differentiation axis rather than within
categories, `t50window()` splits genes by group, orders each group by t50, and
chops it into equal-sized windows named by their mean t50:

```{r}
t50_df <- data.frame(
  gene  = paste0("g", 1:20),
  t50   = seq(1, 40, length.out = 20),
  group = rep("all", 20)
)
windows <- t50window(t50_df, window_number = 4)
windows[[1]]
```

Each column is a window and its name is that window's mean t50, which feeds
`topGO_timeseries()` directly. Note that `floor(n / window_number)` genes go
in each window, so up to `window_number - 1` of the highest-t50 genes are
dropped.

## Plotting trajectories

Given a gene-to-cluster mapping, three related functions draw mean
trajectories per cluster:

- `plot_cluster_line_EBseq()` — one panel per cluster, one line per study,
  optional SD ribbon, plotted against sampled time points
- `plot_cluster_line_interp()` — the same against the interpolated index, and
  can annotate each panel with its cluster's mean t50
- `plot_group_timeline()` — panels *ordered along* the time axis by each
  cluster's median t50

```{r, eval = FALSE}
plot_group_timeline(interp_list, meta_list, gene2cluster,
                    add_t50 = t50_table, time_ref = time_ref)
```

These three are ~260 lines each with pairwise diffs under 90 lines; they want
to be one function with arguments, which is noted in `REVIEW.md`.
`line_scatter_facet()` pairs each cluster's mean with a scatter of its
individual genes.
